---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
# Multi-way-number-partitioning

<!-- badges: start -->
<!-- badges: end -->

## Introduction

The multi-way number partitioning problem tries to partition the elements of a finite set $I$ of integers into $K > 0$ mutually exclusive subsets $I_k$ such that their sum is as close together as possible. More formally: let $S_k = \sum_{i \in I_k}i$, then the goal is to minimize $\max_{k \in K}S_k - \min_{k \in K}S_k$ [ref](https://en.wikipedia.org/wiki/Partition_problem).

## Mixed-integer programming model

One way to solve such a problem is to formulate it as a mixed-integer linear program.

```{r, echo=FALSE}
library(ROI.plugin.glpk)
library(rmpk)
model <- MIPModel(ROI_solver("glpk"))
K <- 4
n_I <- 10
I <- integer(n_I)
```

### Variables

For each element $i$ and $k$ $x_{i,k}$ is 1 if element $i$ is part of set $S_k$.
```{r}
x <- model$add_variable(
  i = 1:n_I,
  k = 1:K,
  type = "integer",
  lb = 0,
  ub = 1
)
```

Then we need three more variables to keep track of the size of the subsets.

```{r}
size <- model$add_variable(k = 1:K)
max_size <- model$add_variable()
min_size <- model$add_variable()
```


### Objective

Minimize the difference between the largest and the smallest set.

```{r}
model$set_objective(max_size - min_size, sense = "min")
```

### Constraints

Each element $i$ needs to part of exactly one subset $S_k$.
```{r}
model$add_constraint(
  sum_expr(x[i, k], k = 1:K) == 1,
  i = 1:n_I
)
```

The following constraints model the correct size of each subset and define the correct values of the minimum and maximum subset values.

```{r}
model$add_constraint(
  sum_expr(x[i, k] * I[i], i = 1:n_I) == size[k],
  k = 1:K
)
model$add_constraint(size[k] >= min_size, k = 1:K)
model$add_constraint(size[k] <= max_size, k = 1:K)
```

Some notes:

* I did test a formulation which orders the subsets by size in or break symmetries. This lead to wors computational results compared to the model above.
* However I do need to run a full computational study with a larger number of samples to make any claims.
* Once I have a larger testset I will also test CBC.
* This needs to be compared with the greedy heuristic implemented by Noam Ross.

### Computational results

We use *GLPK* as a solver and see how far we can go.

#### The implementation

```{r}
library(rmpk)
library(ROI.plugin.glpk)
solve_partitioning_problem <- function(I, K, time_limit_sec = 30) {
  n_I <- length(I)
  model <- MIPModel(ROI_solver("glpk", control = list(verbose = TRUE, presolve = TRUE, tm_limit = time_limit_sec * 1000)))
  x <- model$add_variable(
    i = 1:n_I,
    k = 1:K,
    type = "integer",
    lb = 0,
    ub = 1
  )
  size <- model$add_variable(k = 1:K)
  max_size <- model$add_variable()
  min_size <- model$add_variable()
  model$set_objective(max_size - min_size, sense = "min")
  model$add_constraint(
    sum_expr(x[i, k], k = 1:K) == 1,
    i = 1:n_I
  )
  model$add_constraint(
    sum_expr(x[i, k] * I[i], i = 1:n_I) == size[k],
    k = 1:K
  )
  model$add_constraint(size[k] >= min_size, k = 1:K)
  model$add_constraint(size[k] <= max_size, k = 1:K)
  model$optimize()
  list(model = model, solution = model$get_variable_value(x[i, k]))
}
```

We also need a function score a solution and to display it.

```{r}
suppressPackageStartupMessages(library(dplyr, warn.conflicts = FALSE))
display <- function(result, I) {
  solution <- result$solution
  solution %>% 
    filter(value == 1) %>% 
    group_by(k) %>% 
    summarise(sum = sum(I[i]), n = n())
}
score <- function(result, I) {
  subset_sum <- display(result, I)$sum
  max(subset_sum) - min(subset_sum)
}
```

#### Small example

```{r}
set.seed(42)
K <- 4
n_I <- 10
I <- sample.int(n_I, size = 10000, replace = TRUE, prob = runif(n_I))
I <- sort(as.integer(table(I)), decreasing = TRUE)
```


```{r}
result <- solve_partitioning_problem(I, K)
```

```{r}
stopifnot(result$model$termination_status() == rmpk::TERMINATION_STATUS$SUCCESS)
```

Let's check the score. Note: this should be the same value as the final upper bound in the log file.
```{r}
score(result, I)
```

```{r}
stopifnot(score(result, I) == result$model$objective_value())
```

```{r}
display(result, I)
```

#### Small K, large I

```{r}
set.seed(42)
K <- 4
n_I <- 200
I <- sample.int(n_I, size = 3e6, replace = TRUE, prob = runif(n_I))
I <- sort(as.integer(table(I)), decreasing = TRUE)
```


```{r}
result <- solve_partitioning_problem(I, K)
```

After 30 seconds the score is: `r result$model$objective_value()`

```{r}
display(result, I)
```
